Stochastic Properties of EIP-1559 Basefees

### 2021-05-20 eip1559_arxiv/template.tex
MDPI submission

%\date{September 9, 1985}	% Here you can change the date presented in the paper title
%\date{} 					% Or removing it

\author{ \href{https://orcid.org/0000-0000-0000-0000}{\includegraphics[scale=0.06]{orcid.pdf}\hspace{1mm}Ian C. Moore, PhD}\\
	\texttt{ic3moore@gmail.com} \\
	%% examples of more authortexmaker
	\And
	\href{https://orcid.org/0000-0000-0000-0000}{\includegraphics[scale=0.06]{orcid.pdf}\hspace{1mm}Jagdeep Sidhu, MSc} \\
	Syscoin Core Developer\\
	Blockchain Foundry Inc.\\
	\texttt{jsidhu@blockchainfoundry.co} \\
	%% \AND
	%% Coauthor \\
	%% Affiliation \\
	%% Address \\
	%% \texttt{email} \\
	%% \And
	%% Coauthor \\
	%% Affiliation \\
	%% Address \\
	%% \texttt{email} \\
	%% \And
	%% Coauthor \\
	%% Affiliation \\
	%% Address \\
	%% \texttt{email} \\
}

% Uncomment to remove the date
%\date{}

% Uncomment to override  the `A preprint' in the header
%\renewcommand{\headeright}{Technical Report}
%\renewcommand{\undertitle}{Technical Report}
\renewcommand{\shorttitle}{EIP-1559 Basefees}

%%% Add PDF metadata to help others organize their library
%%% Once the PDF is generated, you can check the metadata with
%%% $ pdfinfo template.pdf
\hypersetup{
pdftitle={Stochastic Properties of EIP-1559 Basefees},
pdfsubject={q-bio.NC, q-bio.QM},
pdfauthor={Ian C. Moore, Jagdeep Sidhu},
pdfkeywords={EIP 1559, Base Fees, Stochastic Processes, Stationarity},
}

@@

\begin{abstract}
EIP-1559 is a new proposed pricing mechanism for the Ethereum protocol developed to bring stability to fluctuating gas prices. To properly understand this as a stochastic process, it is necessary to develop the mathematical foundations to understand under what conditions the base fee gas price outcomes behave as a stationary process, and when it does not. Understanding these mathematical fundamentals is critical to properly engineering a stable system.
\end{abstract}

% keywords can be removed
\keywords{EIP 1559  \and Base Fees  \and Stochastic Processes  \and Stationarity}

\section{Introduction}
The current Ethereum pricing mechanism employs the price auction model where high-value use cases are prioritized over lower ones. The problem with this approach is that as Ethereum has grown more popular, users have difficulty estimating optimal gas fees. In 2018, EIP-1559 was  proposed by Ethereum Founder, Vitalik Buterin, as a major change to Ethereum's transaction fee mechanism, and introduces two different types of fees, which is the base fee and inclusion fee \cite{But19}. The ideas is the base fee is relatively stable while the inclusion fee serves as an additional tip offered to compensate miners. This mechanism allows for the base fee to be adjusted after every block in accordance to demand. Hence, when demand for gas is higher than the target gas price, the block size is adjusted upwards, and when demand for gas is lower than the target gas, the block size is adjusted downwards. Blocks are allowed to grow as large as double the target block size.

In this study, we investigate the stochastic properties of EIP-1559 base fees. As the main motivation behind EIP-1559 is to bring stability to gas fees, we look at the mathematical conditions that this mechanism must have to achieve stationarity. We open up this discussion in Section \ref{section:eip_1559} where we go over the EIP 1559 pricing mechanism. Using these simulated gas demands, we analyze the stationary properties in Section \ref{section:analyze_stationarity}. In Section \ref{section:gas_demand} we simulate the gas demand.  Finally, we close off with our conclusions in Section \ref{section:conclusion}.

@@

\begin{equation}
f_{k} = 1 + \frac{\delta_{k}}{c}
\label{eq:fk} 
\end{equation}
where $\delta_{k} = demand_{k} - Target Gas Fee$ and $c = Target ~/~ BaseFee Max Change$; for the purpose of this study, we assume $\delta_{k}$ to be a random stationary process. However before we do that, we run the \textit{abm1559} simulator \cite{Mon21} to show that this is the case, under the assumption that new users are Poisson distributed.

@@

\textbf{Definition:} A time series ${X_{t}}$ is called strictly stationary if the random vectors $(X_{1} , ... , X_{N} )^T$ and $(X_{1+\tau} , ... , X_{N+\tau} )^T$ have the same joint distribution for all sets of indices ${t_{1} , ... , t_{n}}$ and for all integers $\tau$ and $n > 0$. It is written as

 \begin{equation}
(X_{1} , ... , X_{N} )^T \stackrel{D}{=} (X_{1+\tau} , ... , X_{N+\tau} )^T,
\label{eq:ar1} 
\end{equation}
where $\stackrel{D}{=}$ means equal in distribution.

@@

\begin{equation}
\delta y_{t} = \alpha y_{t-1} + \theta_{1} \delta Y_{t-1}  + ... + \theta_{p} \delta Y_{t-p} + \epsilon_{t},
\label{eq:ar1} 
\end{equation}
the Augmented Dickey Fuller (ADF) test checks the existence of a unit root $\alpha = 1$ (ie, H0: $\alpha = 1$) where the null hypothesis is non-stationary. The unit root in (\ref{eq:ar1}) is characteristic of a time series that makes it non-stationary.

@@

In Section \ref{section:gas_demand} we simulated gas demand  $\delta_{k}$ in (\ref{eq:fk}) and found them to be stationary and normally distributed with $\mu$ = 13641075, and variance $\sigma^2$ = 550980. To simulate gas prices using EIP 1559 we used these parameter estimates to model a year's worth of gas demands and fed it through the system described in (\ref{eq:eip1559}) and (\ref{eq:fk}). A series of these simulations can be seen in Figure \ref{fig:basefee_simulations}.

@@

\subsection{Problem Statement}
\label{section:problem_statement}
For sake of comparison, let's look at stationarity for an AR(1) process. Consider a standard first-order auto-regressive process defined by the recursive equation:

\begin{equation}
x_{t} = \mu + \alpha (x_{t-1} -\mu) + \epsilon_{t-1} ~~~~~~~~ \epsilon_{t} \sim IID~N(0,\sigma^2)
\label{eq:ar1} 
\end{equation}
The above process is stationary when $|\alpha|$<1 so that this has a single root 1/$\alpha$ outside the unit circle.  It can be shown that if $\sigma$>0, then this process has the asymptotic stationary distribution:

\begin{equation}
x_{\infty} \sim N(\mu,\frac{\sigma^2}{1-\alpha^2})
\label{eq:x_infinity} 
\end{equation}
Considering the (state space like) simularities to the AR(1) process to the system of the system of (\ref{eq:eip1559}) and (\ref{eq:fk}); the question is what are the constraints (if any) that time varying process $\delta_{t}$ must meet to ensure stationary outcomes for the basefees.

@@

\begin{equation}
x_{k} = \alpha + \beta_{k}x_{k-1} + \epsilon_{k},
\label{eq:rca1}
\end{equation}
where ${\epsilon_{k}}$ is an independent sequence of random variables with 0 mean and variance $\sigma^2 > 0$; ${\beta_{k}}$ is an independent sequence of random variables with mean $\mu_{\beta}$ and variance $\sigma_{\beta}^2$ of the random coefficient $\beta_{k}$, and variance $\sigma^2$ of the error $\epsilon_{k}$. Given (3) and (4) it is obvious that when $\sigma_{\beta}^2 = 0$, the RCA(1) in (5) becomes a AR series, and becomes a random walk when $\mu_{\beta} = 1$ and $\sigma_{\beta}^2 = 0$, hence non-stationary.

EIP 1559 (1) fits into the framework of a Random Coefficient Autoregressive (RCA) model when $\epsilon$ is negligibly small. This is good news, as it allows us to apply the body of work on RCA processes to this problem. The next question is; under what conditions does an RCA(1) process exhibit stationary behaviour. If we know this, then we can apply this understanding to the EIP 1559 mechanism.

@@

\textbf{Proof:} See Theorem 2 in \cite{Wan03}

\begin{figure}
\centering
\includegraphics[width=5in]{strict_stationarity.png}
\caption{Region of strict stationarity and ergodicity for RCA(1) determined via (\ref{eq:thm}). The annotated red dot represents where our setup is in relation to the region where an RCA(1) process behaves as a strictly stationary process. Based on this we can see that our setup was non-stationary.} 
\label{fig:strict_stationarity}
\end{figure}

@@

The purpose of this study was to understand under what conditions the EIP 1559 pricing mechanism would behave as a stationary process. We believe this as a critical first step to understanding the constraints to engineer stable system. Given the enormity of the Ethereum project, understanding these fundamentals are needed for setting up simulation experiments to help capture all the edge cases that may arise so that the number of updates and patches are minimized prior to release.

It was determined that under the parameter setting used in this study that the outcomes behaved as Brownian motion, which is a non-stationary process. The RCA(1) was used as the mathematical framework to help understand this, as we can go to the literature to help guide our understanding of this kind of process. This conclusion was based on a series of simulations which we tested using the ADF, and Wang's Theorem indicating the region of stationarity for an RCA(1) process.

>>>>

\author{ \href{https://orcid.org/0000-0000-0000-0000}{\includegraphics[scale=0.06]{orcid.pdf}\hspace{1mm}Ian C. Moore, PhD}\\
	Syscoin Researcher\\
	\texttt{imoore@syscoin.org} \\
	\And
	\href{https://orcid.org/0000-0000-0000-0000}{\includegraphics[scale=0.06]{orcid.pdf}\hspace{1mm}Jagdeep Sidhu, MSc} \\
	Syscoin Core Developer\\
	Blockchain Foundry Inc.\\
	\texttt{jsidhu@blockchainfoundry.co} \\
}

\renewcommand{\shorttitle}{EIP-1559 Basefees}

\hypersetup{
pdftitle={Stochastic Properties of EIP-1559 Basefees},
pdfsubject={q-bio.NC, q-bio.QM},
pdfauthor={Ian C. Moore, Jagdeep Sidhu},
pdfkeywords={EIP 1559, Base Fees, Stochastic Processes, Stationarity},
}

@@

\begin{abstract}
EIP-1559 is a new proposed pricing mechanism for the Ethereum protocol developed to mitigate short-term volatility in demand for transactions. To properly understand this as a stochastic process, it is necessary to develop the mathematical foundations to understand under what conditions the base fee gas price outcomes behave as a stationary process, and when it does not. We believe understanding these mathematical fundamentals is critically needed to properly engineering a well-designed system.
\end{abstract}

\keywords{EIP 1559  \and Base Fees  \and Stochastic Processes  \and Stationarity}

\section{Introduction}
The current Ethereum pricing mechanism employs the price auction model where high-value use cases are prioritized over lower ones. The problem with this approach is that as Ethereum has grown more popular, users have difficulty estimating optimal gas fees. In 2018, EIP-1559 was  proposed by Ethereum Founder, Vitalik Buterin, as a major change to Ethereum's transaction fee mechanism, and introduces two different types of fees, which is the base fee and inclusion fee \cite{But19,Rou20}. The idea is the base fee is relatively stable while the inclusion fee serves as an additional tip offered to compensate miners. This mechanism allows for the base fee to be adjusted after every block in accordance to demand. Hence, when demand for gas is higher than the target gas price, the block size is adjusted upwards, and when demand for gas is lower than the target gas, the block size is adjusted downwards. Blocks are allowed to grow as large as double the target block size.

In this study, we investigate the stochastic properties of EIP-1559 base fees. As the main motivation behind EIP-1559 is to address the volatility and uncertainty associated with gas fees, we look at the mathematical conditions that this mechanism must have to achieve stationarity. We open up this discussion in Section \ref{section:eip_1559} where we go over the EIP 1559 pricing mechanism. Using these simulated gas demands, we analyze the stationary properties in Section \ref{section:analyze_stationarity}. In Section \ref{section:gas_demand} we simulate the gas demand.  Finally, we close off with our conclusions in Section \ref{section:conclusion}.

@@

\begin{equation}
f_{k} = 1 + \frac{\delta_{k}}{c}
\label{eq:fk} 
\end{equation}
where $\delta_{k} = demand_{k} - Target~Gas~Fee$ and $c = Target ~/~ Base~Fee~Max~Change$; for the purpose of this study, we applied setup from Table 1, and assumed $\delta_{k}$ to be a random stationary process. However before we do that, we run the \textit{abm1559} simulator \cite{Mon21} to show that this is the case, under the assumption that new users are Poisson distributed.

@@

\textbf{Definition:} A time series $\{x_{t}\}$ is said to be strictly stationary if random vectors $(x_{t_{1}} , ... , x_{t_{n}} )^T $ and $(x_{t_{1}+\tau} , ... , x_{t_{n}+\tau} )^T$ have the same joint distribution for all $\{t \in \mathbb{N}^{0}\}$ and $n,\tau \in \mathbb{N}^{+}$. This is defined as:
\begin{equation}
(x_{t_{1}} , ... , x_{t_{n}})^T \stackrel{d}{=} (x_{t_{1}+\tau} , ... , x_{t_{n}+\tau} )^T,
\label{eq:ar1} 
\end{equation}
where $\stackrel{d}{=}$ means equivalence in distribution.

@@

\begin{equation}
\Delta y_{t} = \alpha y_{t-1} + \theta_{1} \Delta y_{t-1}  + ... + \theta_{p} \Delta y_{t-p} + \epsilon_{t},
\label{eq:ar1} 
\end{equation}
the Augmented Dickey Fuller (ADF) test checks the existence of a unit root $\alpha = 1$ (ie, H0: $\alpha = 1$) where the null hypothesis is non-stationary. The unit root in (\ref{eq:ar1}) is characteristic of a time series that makes it non-stationary.

@@

In Section \ref{section:gas_demand} we simulated gas demand  $\delta_{k}$ in (\ref{eq:fk}) and found them to be stationary and normally distributed with $\mu$ = 1.36e7, and variance $\sigma^2$ = 5.51e5. To simulate gas prices using EIP 1559 we used these parameter estimates to model a year's worth of gas demands and fed it through the system described in (\ref{eq:eip1559}) and (\ref{eq:fk}). A series of these simulations can be seen in Figure \ref{fig:basefee_simulations}.

@@

\subsection{Problem Statement}
\label{section:problem_statement}
For sake of comparison, let's look at the stationarity for a standard first-order auto-regressive process  (ie, AR(1)) defined as:

\begin{equation}
x_{t} = \mu + \alpha (x_{t-1} -\mu) + \epsilon_{t-1} ~~~~~~~~ \epsilon_{t} \sim IID~N(0,\sigma^2)
\label{eq:ar1} 
\end{equation}
The above process is stationary when $|\alpha|$<1, hence has a single root 1/$\alpha$ outside the unit circle. The asymptotic stationary distribution can be shown to be:
\begin{equation}
x_{\infty} \sim N(\mu,\frac{\sigma^2}{1-\alpha^2}),
\label{eq:x_infinity} 
\end{equation}
when $\sigma$>0. Considering the (state space like) simularities to the AR(1) process to the system of the system of (\ref{eq:eip1559}) and (\ref{eq:fk}); the question is what are the constraints (if any) that time varying process $\delta_{t}$ must meet to ensure stationary outcomes for the basefees.

@@

\begin{equation}
x_{t} = \alpha + \beta_{t}x_{t-1} + \epsilon_{t},
\label{eq:rca1}
\end{equation}
where ${\epsilon_{t}}$ is an independent sequence of random variables with 0 mean and variance $\sigma^2 > 0$; ${\beta_{t}}$ is an independent sequence of random variables with mean $\mu_{\beta}$ and variance $\sigma_{\beta}^2$ of the random coefficient $\beta_{t}$, and variance $\sigma^2$ of the error $\epsilon_{t}$. Given (3) and (4) it is obvious that when $\sigma_{\beta}^2 = 0$, the RCA(1) in (5) becomes a AR series, and becomes a random walk when $\mu_{\beta} = 1$ and $\sigma_{\beta}^2 = 0$, hence non-stationary.

EIP 1559 (1) fits into the framework of a Random Coefficient Autoregressive (RCA) model when $\epsilon$ is negligibly small. This is good news, as it allows us to apply the body of work on RCA processes to this problem. The next question is; under what conditions does an RCA(1) process exhibit stationary behaviour. If we know this, then we can apply this understanding to the EIP 1559 mechanism. To help with our understanding regarding this, we apply Wang's Theorem \cite{Wan03}; see Figure \ref{fig:strict_stationarity}.

@@

\textbf{Proof:} See Theorem 2 in \cite{Wan03}.

\begin{figure}
\centering
\includegraphics[width=5in]{strict_stationarity.png}
\caption{Region of strict stationarity and ergodicity for RCA(1) determined via (\ref{eq:thm}). The annotated red dot represents where our setup (Table 1) is in relation to the region where an RCA(1) process behaves as a strictly stationary process. Based on this we can see that our setup was non-stationary.} 
\label{fig:strict_stationarity}
\end{figure}

@@

The purpose of this study was to understand under what conditions the EIP 1559 pricing mechanism would behave as a stationary process. We believe this as a critical first step to understanding the constraints to engineer stable system. Given the enormity of the Ethereum project, understanding these fundamentals are needed for setting up well-designed simulation experiments to help capture all the edge cases that may arise so that the number of updates and patches are minimized prior to release.

It was determined that under the parameter setting used in this study that the outcomes behaved as a random walk, which is a non-stationary process. The RCA(1) was used as the mathematical framework to help understand this, as we can go to the literature to help guide our understanding of this kind of process. This conclusion was based on a series of simulations which we tested using the ADF, and Wang's Theorem indicating the region of stationarity for an RCA(1) process.
